Popularity Adjusted Block Models are Generalized Random Dot Product Graphs

### 2021-08-29 summary.tex
syncing stuff again

Using \(A\) instead of \(P\) introduces error, which converges to \(0\)
almost surely. We use this fact to motivate Orthogonal Spectral Clustering 
(Alg. 1). 
In order to show that OSC is consistent, 
we introduce the sparsity factor $\rho_n \in (0, 1]$, 
which can be characterized as the average expected degree of the vertices 
divided by the number of vertices. 
The situation is the same as in \cite{307cbeb9b1be48299388437423d94bf1} (PABM) 
and \cite{rubindelanchy2017statistical} (GRDPG) in which the popularity vectors 
are scaled as $\rho_n^{1/2} \lambda^{(k \ell)}$. 
For ease of notation, we restrict our results to the case in which all 
nonzero popularity parameters are greater than some positive minimum value.

\begin{theorem}
\label{theorem4}
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix from OSC
(Alg. 1). Then for all pairs $(i, j)$ belonging to different communities
and sparsity factor satisfying $n \rho_n = \omega(\log^{4}{n})$, we have
\begin{equation} \label{eq:thm4}
\max_{i, j} |n \hat{v}_i^\top \hat{v}_j| =
O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
\end{equation}
with high probability. Eq.~\eqref{eq:thm4} implies that for all pairs $(i, j)$ in different communities,
we have $\hat{B}_{ij} \stackrel{a.s.}{\to} 0$ where the convergence is
uniform over all pairs $(i,j)$. 
\end{theorem}

@@

We also define two types of parameter estimation error. 
First, we estimate the popularity vectors directly and compute the RMSE. 
In this case, the ``true'' popularity vectors are derived from 
taking the SVD of each edge probability block $P^{(k \ell)}$ 
to avoid the unidentifiable multiplicative constants, i.e., 
\[\text{RMSE}(\{\hat{\lambda}^{(k \ell)}\}, \{\tilde{\lambda}^{(k \ell)}\}) = 
\Bigl(\frac{1}{n} \sum_{k < l} \min_{s = \pm 1} 
\|s \hat{\lambda}^{(k \ell)} - \tilde{\lambda}^{(k \ell)} \|_2^2\Bigr)^{1/2}.\]
We can also avoid the unidentifiable multiplicative constant more directly 
by reconstructing each 
$\hat{P}^{(k \ell)} = \hat{\lambda}^{(k \ell)} (\hat{\lambda}^{(\ell k)})^\top$, 
which we use to define another parameter estimation error, i.e., 
\[\text{RMSE}(\hat{P}, P) =
\sum_{k,l} \sqrt{\frac{1}{n_k n_\ell} \|P^{(k \ell)} - \hat{P}^{(k \ell)}\|_F^2}.\]

>>>>

Using \(A\) instead of \(P\) introduces error, which converges to \(0\)
almost surely. We use this fact to motivate Orthogonal Spectral Clustering 
(Alg. 1). 
In order to show that OSC is consistent, 
we introduce the sparsity factor $\rho_n \in (0, 1]$, 
which can be characterized as the average expected degree of the vertices 
divided by the number of vertices. 
The treatment is the same as in \cite{307cbeb9b1be48299388437423d94bf1} (PABM) 
and \cite{rubindelanchy2017statistical} (GRDPG) in which the popularity vectors 
are scaled as $\rho_n^{1/2} \lambda^{(k \ell)}$. 
This results in the edge probability matrix being scaled as $\rho_n P$. 
Since $\rho_n$ is unidentifiable, we use the convention that the popularity parameters are drawn such that $\|X\|_F = O(\sqrt{n})$ and require the $L^2$ norm of each row is bounded away from 0 and 1.

\begin{theorem}
\label{theorem4}
Let $\hat{B}$ with entries $\hat{B}_{ij}$ be the affinity matrix from OSC
(Alg. 1). Then for all pairs $(i, j)$ belonging to different communities
and sparsity factor satisfying $n \rho_n = \omega(\log^{4}{n})$, we have
\begin{equation} \label{eq:thm4}
\max_{i, j} |n \hat{v}_i^\top \hat{v}_j| =
O\Big( \frac{\log n}{\sqrt{n \rho_n}} \Big)
\end{equation}
with high probability. 
\end{theorem}

Eq.~\eqref{eq:thm4} implies that for all pairs $(i, j)$ in different communities,
we have uniform convergence $\hat{B}_{ij} \stackrel{a.s.}{\to} 0$. 
This allows us to identify the communities perfectly from $\hat{B}$ for large $n$. 
For example, if we use edge thresholding to partition $\hat{B}$, 
Theorem \ref{theorem4} implies that for any $\epsilon > 0$, 
the number of edges of $\hat{B}$ between vertices of different communities 
that are larger than $\epsilon$ converges to zero with probability 1.

@@

For parameter estimation, because the popularity parameters $\{\lambda_{ik}\}$ are unidentifiable, we instead estimate the edge probabilities $p_{ij} = \lambda_{i z_j} \lambda_{j, z_i}$. The parameter estimation error is the root mean square error between the true and estimated edge probabilities, which is equivalent to the Frobenius norm of $P$:

$$\mathrm{RMSE}(\hat{P}, P) = \|\hat{P} - P\|_F.$$
